United Monoids

### 2021-10-04 paper/main.tex
Proofreading

  \item We introduce \emph{united monoids} as an algebraic structure that
  captures the essence of the various studied interpretations of trees
  (\S\ref{sec-united}). We also characterise united monoids with a minimal set
  of axioms.
\end{itemize}

@@

When working with simplicial sets, we will extend the requirements on
interpreting trees with \hcode{fold} by adding the distributivity requirement:

\begin{itemize}
    \item Associativity, inherited from~\S\ref{sec-tree}:
          $(a \dia b) \dia c = a \dia (b \dia c)$,
          $(a \arr b) \arr c = a \arr (b \arr c)$;
    \item Commutativity, inherited from~\S\ref{sec-set}:
          $a \dia b = b \dia a$;
    \item Idempotence, inherited from~\S\ref{sec-set}:
          $a \dia a = a$;
    \item Distributivity (\textbf{\color{darkblue}new!}):
          $a \arr (b \dia c) = (a \arr b) \dia (a \arr c)$,
          $(a \dia b) \arr c = (a \arr c) \dia (b \arr c)$.
\end{itemize}

\noindent
The new requirement is motivated by the desire to make the trees
\code{("a" $\,\hdia\,$ "b") $\,\harr\,$ "c"} and
\code{("a" $\,\harr\,$ "b") $\,\hdia\,$ ("a" $\,\harr\,$ "c")} mean the same
simplicial set (the rightmost one above).

@@

\noindent
Here the inner non-empty lists correspond to simplices, with their vertices
listed in the connection order. For example, the simplicial set corresponding to
the tree \code{"a" $\,\harr\,$ "b"} will contain three lists: \code{[a]},
\code{[b]} and \code{[a,b]}.

@@

\noindent
Note that \hcode{SimplicialSet} is a rather naive representation for simplicial
sets. In particular, an $n$-simplex contains $O(2^n)$ sub-simplices of
lower dimension, and all of them will be included in the outer \hcode{Set}. It
is possible to improve the representation by storing only maximal simplices, or
by using suffix trees~\cite{weiner_suffix_trees}. The \hcode{Tree}-based
representation of simplicial sets is free from this problem: an $n$-simplex can
be described by an expression of size $O(n)$, as evidenced by the function
\hcode{simplex} in Listing~\ref{lst-simplicial-set}.

Another advantage of the \hcode{Tree}-based representation is that it is free
from internal invariants, i.e. any value of \hcode{Tree Bool a} describes a
valid simplicial set. \hcode{SimplicialSet} does not satisfy this property; for
example, the set containing lists \code{[a]} and \code{[a,b]} is not a
simplicial set, because the edge $\textit{ab}$ appears in the set without its
vertex $\textit{b}$.

@@

\begin{lstlisting}
type Graph a = (Set a, Set (a, a)) -- A set of vertices and a set of edges
\end{lstlisting}

@@

\noindent
This law is unusual but it is not new: it was introduced as an axiom of
\emph{algebraic graphs} in~\cite{mokhov_alga}. In fact, by adding this
requirement to those used when interpreting simplicial sets
in~\S\ref{sec-simplicial-set}, we get exactly the algebra of non-empty graphs
from~\cite{mokhov_alga}. It is therefore not surprising that we can reuse a lot
of more general functions operating on trees and simplicial sets when
implementing a part of the API of algebraic graphs -- see
Listing~\ref{lst-graph}, where \hcode{hasEdge} is the only function that we did
not reuse.

@@

\noindent
All we need to do is adapt \hcode{vertex}, \hcode{overlay} and \hcode{connect}
to ``truncated'' simplicial sets. The ``truncated'' variants of \hcode{vertex}
and \hcode{overlay} are straightforward:

@@

\noindent
As for the edge set, one can mechanically check that both sides of the
decomposition requirement evaluate to the following symmetric expression.

@@

hasEdge :: Ord a => a -> a -> TGraph a -> Bool --
hasEdge x y = maybe False (elem (x,y) . edgeSet) . induce (\a -> a == x || a == y)
!\rule[2mm]{\textwidth}{0.4pt}!
\end{lstlisting}

@@

\noindent
One can notice that this rule is reminiscent of the relaxation step in various
shortest-path finding algorithms, where a distance between two vertices $a$ and
$c$ is updated after discovering a transitive route via a vertex $b$ (recall
that in semirings~\mul~is used for sequential composition of distances).

@@

Armed with the notation from the previous section~\S\ref{sec-preorder}, we can
now use our little tree language to describe graphs whose edges are labelled
with values from an arbitrary semiring \hcode{s}. For example, the expression
$\text{\code{"a"}} \xrightarrow{\one} \text{\code{"b"}}\ \dia\ \text{\code{"b"}} \xrightarrow{\two} \text{\code{"c"}}$
will correspond to the graph below.

@@

\noindent
To translate the shortcut notation to Haskell, we define the following two
operators:

\begin{lstlisting}
(-<) :: Tree s a -> s -> (Tree s a, s)
x -< s = (x, s)
!\vspace{-1mm}!
(>-) :: (Tree s a, s) -> Tree s a -> Tree s a
(x, s) >- y = node s x y
\end{lstlisting}

\noindent
This allows us to write the above example as
\code{"a"}~\code{-<1>-}~\code{"b"}~\hdia~\code{"b"}~\code{-<2>-}~\code{"c"}.

\newpage
\noindent
We will interpret such tree expressions into the following data type of
edge-labelled graphs, which maps vertices \hcode{a} to a collection of
\hcode{s}-labelled outgoing edges.

@@

overlay :: (Ord a, Semiring s) => LGraph s a -> LGraph s a -> LGraph s a
overlay = Map.unionWith (Map.unionWith (!\h{\hadd}!)) -- Combine parallel edge labels with !\h{\hadd}!
\end{lstlisting}

@@

\vspace{-5mm}
\begin{equation*}
\begin{array}{rcll}
a & = & \one\ \mul\ a & \text{(unit of $\mul$)}\\
 & = & \zero\ \mul\ a & \text{(we postulate \zero = \one)}\\
 & = & \zero & \text{(annihilating \zero)}
\end{array}
\end{equation*}
\vspace{-3mm}

@@

\vspace{-5mm}
\begin{equation*}
\begin{array}{rcll}
a & = & 1\ \wedge\ a & \text{(unit of $\wedge$)}\\
 & = & 1\ \wedge\ (0\  \vee\  a) & \text{(unit of $\vee$)}\\
 & = & 0\ \wedge\ (0\  \vee\  a) & \text{(we postulate 0 = 1)}\\
 & = & 0 & \text{(absorption axiom)}
\end{array}
\end{equation*}
\vspace{-3mm}

@@

Consider two monoids $(S, +, 0)$ and $(S, \cdot, 1)$, such that $+$ is
commutative and $\cdot$ distributes over $+$. We call these monoids
\emph{united} if $0 = 1$. To avoid confusion with semirings and lattices, we
will use $\varepsilon$ to denote the unit element of both monoids, that is,
$a + \varepsilon = a \cdot \varepsilon = a$. Note: we will often omit the
operator $\cdot$ and write simply $ab$ instead of $a \cdot b$, which is a usual
convention. We will further refer to $\varepsilon$ as \emph{empty}, the
operation~$+$ as \emph{overlay}, and the operation~$\cdot$ as \emph{connect}.

@@

The next property of united monoids is more unusual:
$ab = ab + a = ab + b = ab + a + b$. We call these equalities the \emph{containment laws}:
intuitively, when you connect $a$ and $b$, the constituent parts are contained
in the result $ab$. Let us prove containment:

\vspace{-5mm}
\begin{equation*}
\begin{array}{rcll}
ab\ +\ a & = & ab\ +\ a\varepsilon & \text{(unit of $\cdot$)}\\
 & = & a(b\ +\ \varepsilon) & \text{(distributivity)}\\
 & = & ab & \text{(unit of $+$)}\\
\end{array}
\end{equation*}
\vspace{-3mm}

@@

Surprisingly, the containment law $ab = ab + a$ is equivalent to $0 = 1$, i.e.,
the latter can be proved from the former:

@@

\noindent
This means that united monoids can be equivalently characterised by the
containment laws, which makes it possible to talk about
\emph{united semigroups}. The term ``united semigroup'' may sound somewhat
nonsensically (because semigroups have no ``units''), however, note that if they
secretly had unit elements, they would have to coincide.

@@

% \noindent
% Finally, let us prove another unusual property of united monoids: non-empty
% elements of $S$ can have no inverses. More precisely:

@@

The containment laws should have reminded you about simplicial sets since
the latter are closed in terms of containment. For example, a filled-in triangle
contains its edges and vertices, and it cannot appear in a simplicial set
without any of them. This property can be expressed algebraically as:
$abc = abc + ab + ac + bc + a + b + c$. Interestingly, this ``3D'' containment
law follows from the ``2D'' version for united monoids:

\vspace{-5mm}
\begin{equation*}
\begin{array}{rcll}
abc & = & (ab + a + b)c & \text{(containment)}\\
 & = & (ab)c + ac + bc & \text{(distributivity)}\\
 & = & (abc + ab + c) + (ac + a) + (bc + b) & \text{(containment)}\\
 & = & abc + ab + ac + bc + a + b + c & \text{(commutativity)}
\end{array}
\end{equation*}
\vspace{-3mm}

\noindent
We can similarly prove higher-dimensional versions of the containment law; they
all follow from the basic law $ab = ab + a$, or, alternatively, from $0 = 1$.

@@

It is worth remarking on the two remaining cases: sets and edge-labelled graphs.
When working with sets, we had only one operator \hdia, which was caused by the
fact that we used a trivial semiring with \zero~$=$~\one. One can think of this
case as \hdia~being a monoid united with itself.

The case of edge-labelled graphs is more interesting. There we had as many
operations as there were elements in the edge label semiring~\hcode{s}. It turns
out that every operation $\xrightarrow{x}$ where $x$ is non-zero is united with
\hdia. In other words, edge-labelled graphs are a
\emph{semiring of united monoids}.

@@

\begin{itemize}
    \item $(\mathbb{Z}^{\ge 0}, \max, 0)$: the execution time of programs $a$
    and $b$ composed in parallel is defined to be the maximum of their execution
    times:
    \[
    \textit{time}(a\ \text{\code{|}}\ b) = \max(\textit{time}(a),\ \textit{time}(b))
    \]

    \item $(\mathbb{Z}^{\ge 0}, +, 0)$: the execution time of programs $a$ and
    $b$ composed in sequence is defined to be the sum of their execution times:
    \[
    \textit{time}(a\ \text{\code{;}}\ b) = \textit{time}(a) + \textit{time}(b)
    \]
\end{itemize}

@@

Interestingly, parallel and sequential composition of programs also forms a
united monoid, where the empty program is the unit. One can therefore call the
cost model function $\textit{time}$ a \emph{united monoid homomorphism}.
A related example can be found in~\cite{beaumont2017concepts}, where
asynchronous circuit specifications are composed in parallel and sequentially,
using graphs labelled with Boolean predicates.

% What about tries?
% Equivalence classes?
% Languages

\newpage
\section{Related work}\label{sec-related-work}
This paper continues the research on \emph{algebraic graphs} started
in~\cite{mokhov_alga}. Since their introduction, algebraic graphs have been
implemented in several languages (including Agda, Fsharp, PureScript, Scala,
TypeScript, and R) and found application in industry, e.g. in GitHub's static
code analysis project Semantic~\cite{semantic_paper}\cite{semantic_repo}.
In this paper we further distill the essence of the algebraic approach to
working with graphs by simplifying the graph data type from four to just two
constructors, i.e., switching to using binary trees parameterised by a semiring.
One of the main limitations of the work on algebraic graphs was the lack of
edge labels and this paper successfully addresses it.

Other popular approaches for representing graphs in functional programming
languages can be found in~\cite{1995_king_graphs}\cite{1994_launchbury_st}
(adjacency arrays, available from the \textsf{containers} library)
and~\cite{2001_erwig_inductive} (so-called \emph{inductive graphs} where a graph
is decomposed into a \emph{context}, i.e., a node with its neighbourhood, and
the rest of the graph).

Using semirings as a general framework for solving problems on graphs is, of
course, not new. Dolan's ``Fun with Semirings''~\cite{2013_semirings_dolan} is a
notable example from the functional programming community, which motivated the
author to apply semirings in the context of algebraic graphs. A more recent
example~\cite{weighted_search} uses semirings for generalising algorithms for
weighted search.

% This paper does not discuss any algorithms

Jeremy Gibbons defines operations \code{above} and \code{beside} for composing
directed acyclic graphs~\cite{1995_gibbons_algebra}. They both have the same
unit (the empty graph) but the operation \code{beside} is defined only for
graphs of matching types. One can therefore consider this as an example of a
\emph{partial united monoid}. Another similar example is~\cite{typed_matrices},
where compatible matrices can be composed vertically or horizontally with the
empty matrix being the common unit.

Various flavours of parallel and sequential composition often form united
monoids. In a paper on Concurrent Kleene Algebra~\cite{hoare2011concurrent},
the authors use the term ``bimonoid'' to refer to such structures (without
investigating them in detail). In category theory, ``bimonoid'' is used for an
unrelated concept (a structure with a monoid and a
comonoid~\cite{porst_bimonoids}), hence our decision to use the term
``united monoids'', which appears to be unused and also more specific.

Rivas and Jaskelioff discuss various notions of computation as monoids in the
category of endofunctors~\cite{rivas_jaskelioff_2017}. Two of these monoids,
namely \hcode{Applicative} and \hcode{Monad}, turn out to have the same unit
\hcode{pure = return}, and they can therefore be considered \emph{united monoids
in the category of endofunctors}. Investigating categorical equivalents of
the presented ideas remains an interesting direction of future work.

>>>>

  \item We introduce \emph{united monoids} as an algebraic structure that
  captures the essence of the various studied interpretations of trees
  (\S\ref{sec-united}). We also characterise united monoids with a minimal set
  of axioms and prove a few basic theorems.
\end{itemize}

@@

When working with simplicial sets, we will extend the set of requirements on
interpreting trees by adding the \emph{containment} (see~\S\ref{sec-united}) and
distributivity requirements:

\begin{itemize}
    \item Associativity, inherited from~\S\ref{sec-tree}:
          $(a \dia b) \dia c = a \dia (b \dia c)$,
          $(a \arr b) \arr c = a \arr (b \arr c)$;
    \item Commutativity and idempotence, inherited from~\S\ref{sec-set}:
          $a \dia b = b \dia a$ and $a \dia a = a$;
    \item Containment (\textbf{\color{darkblue}new!}):
          $a \arr b = (a \arr b) \dia a \dia b$ (a simplex \emph{contains} its sub-simplices);
    \item Distributivity (\textbf{\color{darkblue}new!}):
          $a \arr (b \dia c) = (a \arr b) \dia (a \arr c)$,
          $(a \dia b) \arr c = (a \arr c) \dia (b \arr c)$.
\end{itemize}

\noindent
Adding distributivity is motivated by the desire to make the trees
\code{("a" $\,\hdia\,$ "b") $\,\harr\,$ "c"} and
\code{("a" $\,\harr\,$ "b") $\,\hdia\,$ ("a" $\,\harr\,$ "c")} mean the same
simplicial set (the rightmost one above).

@@

\noindent
Here the inner non-empty lists correspond to simplices, with their vertices
listed in the connection order. For example, the simplicial set corresponding to
the tree \code{"a" $\,\harr\,$ "b"} will contain three lists, one for each
contained simplex: \code{[a]}, \code{[b]} and \code{[a,b]}.

@@

\noindent
Note that \hcode{SimplicialSet} is a rather naive representation for simplicial
sets. In particular, an $n$-simplex contains $O(2^n)$ sub-simplices of
lower dimension, and all of them will be included in the outer \hcode{Set}. One
can improve \hcode{SimplicialSet} by detecting and storing only maximal
simplices or by switching to suffix trees~\cite{weiner_suffix_trees}. Thanks to
the containment requirement, the \hcode{Tree}-based representation of simplicial
sets makes it easy to define an $n$-simplex with a compact $O(n)$-size
expression -- see the function \hcode{simplex} in Listing~\ref{lst-simplicial-set}.

A key advantage of the \hcode{Tree}-based representation is that it is free from
any internal invariants, i.e. any value of \hcode{Tree Bool a} describes a
(valid) simplicial set. \hcode{SimplicialSet} does not satisfy this property;
for example, the set with two lists \code{[a]} and \code{[a,b]} is not a
simplicial set, because the edge $\textit{ab}$ appears in the set without its
vertex $\textit{b}$.

@@

\begin{lstlisting}
type Graph a = (Set a, Set (a, a)) -- The classic pair !$(V, E)$! with invariant !$E\subseteq V\times V$!
\end{lstlisting}

@@

\noindent
This law is unusual but it is not new: it was introduced as an axiom of
\emph{algebraic graphs} in~\cite{mokhov_alga}. In fact, by adding this
requirement to those used when interpreting simplicial sets
in~\S\ref{sec-simplicial-set}, we get exactly the algebra of non-empty graphs
from~\cite{mokhov_alga}. It is therefore not surprising that we can reuse a lot
of more general functions operating on trees and simplicial sets when
implementing a part of the API of algebraic graphs -- see
Listing~\ref{lst-graph}, where \hcode{hasEdge} is the only new function that we
needed to implement.

@@

\noindent
All we need to do is adapt \hcode{vertex}, \hcode{overlay} and \hcode{connect}
to ``truncated'' simplicial sets. The truncated variants of \hcode{vertex} and
\hcode{overlay} are straightforward:

@@

\noindent
As for the edge set, one can mechanically check that both sides of the
decomposition law evaluate to the following expression when
\code{($\hdia$)}~$=$~\hcode{overlay} and \code{($\harr$)}~$=$~\hcode{connect}.

@@

hasEdge :: Ord a => a -> a -> TGraph a -> Bool
hasEdge x y = maybe False (elem (x,y) . edgeSet) . induce (\a -> a == x || a == y)
!\rule[2mm]{\textwidth}{0.4pt}!
\end{lstlisting}

@@

\noindent
One can notice that this rule is reminiscent of the relaxation step in various
shortest-path finding algorithms, where a distance between two vertices $a$ and
$c$ is updated after discovering a transitive route via a vertex $b$ (recall
that in semirings~\mul~is used for sequential composition of generalised
distances).

% We need parallel composition too!

@@

Armed with the notation from the previous section~\S\ref{sec-preorder}, we can
use our tree language to describe graphs whose edges are labelled with values
from an arbitrary semiring. For example, the expression
$\text{\code{"a"}} \xrightarrow{\one} \text{\code{"b"}}\ \dia\ \text{\code{"b"}} \xrightarrow{\two} \text{\code{"c"}}$
corresponds to the graph below.

@@

\noindent
In Haskell, we can encode this as
\code{"a"}~\code{-<1>-}~\code{"b"}~\hdia~\code{"b"}~\code{-<2>-}~\code{"c"}
by defining a pseudo-ternary operator via two auxiliary binary operators
\code{-<} and \code{>-}:

\begin{lstlisting}
(-<) :: Tree s a -> s -> (Tree s a, s)  !$\hspace{8mm}\vrule\hspace{8mm}$!(>-) :: (Tree s a, s) -> Tree s a -> Tree s a
x -< s = (x, s)                         !$\hspace{39.66mm}\vrule\hspace{8mm}$!(x, s) >- y = node s x y
\end{lstlisting}

\noindent
To give meaning to parallel composition of edges, we add a new folding requirement:

\begin{itemize}
    \item Parallel composition of edge labels:
    $a \xrightarrow{x} b\ \dia\ a \xrightarrow{y} b = a \xrightarrow{x \add y} b$.
\end{itemize}

\noindent
Note that containment~(\S\ref{sec-simplicial-set}) is just a special case of
this new, more general requirement:

\vspace{-5mm}
\begin{equation*}
a \arr b\ \dia\ a\ \dia\ b = a \xrightarrow{\one} b\ \dia\ a \xrightarrow{\zero} b = a \xrightarrow{\one\ \raisebox{0.15mm}{\add}\ \zero} b = a \xrightarrow{\one} b = a \arr b.
\end{equation*}
\vspace{-5mm}

\noindent
With the parallel composition rule in place, we can interpret \hcode{Tree}
expressions into the following ``flat'' representation of edge-labelled graphs,
which maps vertices \hcode{a} to a collection of \hcode{s}-labelled outgoing
edges.

% Can't use Set (a, s, a) for edges: parallel composition can't be easily enforced

@@

overlay :: (Ord a, Semiring s) => LGraph s a -> LGraph s a -> LGraph s a
overlay = Map.unionWith (Map.unionWith (!\h{\hadd}!)) -- Parallel composition of edge labels
\end{lstlisting}

@@

\vspace{-5mm}
\begin{equation*}
\begin{array}{rcll}
a & = & \one\ \mul\ a & \text{(unit of $\mul$)}\\
 & = & \zero\ \mul\ a & \text{(we postulate \zero = \one)}\\
 & = & \zero & \text{(zero of $\mul$)}
\end{array}
\end{equation*}
\vspace{-3mm}

@@

\vspace{-5mm}
\begin{equation*}
\begin{array}{rcll}
a & = & 1\ \wedge\ a & \text{(unit of $\wedge$)}\\
 & = & 1\ \wedge\ (0\  \vee\  a) & \text{(unit of $\vee$)}\\
 & = & 0\ \wedge\ (0\  \vee\  a) & \text{(we postulate 0 = 1)}\\
 & = & 0 & \text{(absorption)}
\end{array}
\end{equation*}
\vspace{-3mm}

@@

Consider two monoids $(S, +, 0)$ and $(S, \cdot, 1)$, such that $+$ is
commutative and $\cdot$ distributes over $+$. We call these monoids
\emph{united} if $0 = 1$. To avoid confusion with semirings and lattices, we
will use $\varepsilon$ to denote the unit element of both monoids, that is,
$a + \varepsilon = a \cdot \varepsilon = a$. Note: we will often omit the
operator $\cdot$ and write simply $\textit{ab}$ instead of $a \cdot b$, which is
a usual convention. We will further refer to $\varepsilon$ as \emph{empty}, the
operation~$+$ as \emph{overlay}, and the operation~$\cdot$ as \emph{connect}.

@@

The next property of is more unusual:
$\textit{ab} = \textit{ab} + a = \textit{ab} + b = \textit{ab} + a + b$. We
call these equalities the \emph{containment laws}: intuitively, when you
connect $a$ and $b$, the constituent parts are contained in the result
$\textit{ab}$. Let us prove containment:

\vspace{-5mm}
\begin{equation*}
\begin{array}{rcll}
\textit{ab}\ +\ a & = & \textit{ab}\ +\ a\varepsilon & \text{(unit of $\cdot$)}\\
 & = & a(b\ +\ \varepsilon) & \text{(distributivity)}\\
 & = & \textit{ab} & \text{(unit of $+$)}\\
\end{array}
\end{equation*}
\vspace{-3mm}

@@

Surprisingly, the containment law $\textit{ab} = \textit{ab} + a$ is equivalent
to $0 = 1$, i.e., the latter can be proved from the former:

@@

\noindent
This means that united monoids can be equivalently characterised by the
containment laws, which makes it possible to also talk about
\emph{united semigroups}. The term ``united semigroup'' may seem somewhat
nonsensical (semigroups have no ``units'' after all), however, note that if they
secretly had unit elements, they would have to coincide.

@@

\noindent
Finally, let us prove another unusual property of united monoids: overlay and
connect can have no inverses with respect to $\varepsilon$. That is: if
$a + b = \varepsilon$ or $\textit{ab} = \varepsilon$ then $a = b = \varepsilon$.

\vspace{-5mm}
\begin{equation*}
\begin{array}{rcllcrcll}
a & = & a + \varepsilon & \text{(unit of $+$)} & \vrule & a & = & a + \varepsilon & \text{(unit of $+$)}\\
 & = & a + a + b & \text{(assumption $a + b = \varepsilon$)}& \vrule &  & = & a + \textit{ab} & \text{(assumption $\textit{ab} = \varepsilon$)} \\
 & = & a + b & \text{(idempotence of $+$)}& \vrule & & = & \textit{ab} & \text{(containment)} \\
 & = & \varepsilon & \text{(assumption $a + b = \varepsilon$)}& \vrule & & = & \varepsilon & \text{(assumption $\textit{ab} = \varepsilon$)}
\end{array}
\end{equation*}
\vspace{-3mm}

\noindent
Therefore, it is not possible to extend united monoids to ring- or field-like
structures with inverses.

% It is time to look at some examples of united monoids.

@@

The containment laws should have reminded you about simplicial sets since
the latter are closed in terms of containment. For example, a filled-in triangle
contains its edges and vertices, and it cannot appear in a simplicial set
without any of them. This property can be expressed algebraically as:
$\textit{abc} = \textit{abc} + \textit{ab} + \textit{ac} + \textit{bc} + a + b + c$.
Interestingly, this ``3D'' containment law follows from the ``2D'' version for
united monoids:

\vspace{-5mm}
\begin{equation*}
\begin{array}{rcll}
\textit{abc} & = & (\textit{ab} + a + b)c & \text{(containment)}\\
 & = & (\textit{ab})c + \textit{ac} + \textit{bc} & \text{(distributivity)}\\
 & = & (\textit{abc} + \textit{ab} + c) + (\textit{ac} + a) + (\textit{bc} + b) & \text{(containment)}\\
 & = & \textit{abc} + \textit{ab} + \textit{ac} + \textit{bc} + a + b + c & \text{(commutativity)}
\end{array}
\end{equation*}
\vspace{-3mm}

\noindent
We can similarly prove higher-dimensional versions of the containment law; they
all follow from the basic law $\textit{ab} = \textit{ab} + a$, or,
alternatively, from $0 = 1$.

@@

It is worth remarking on the two remaining cases: sets and edge-labelled graphs.
When working with sets, we had only one operator \hdia, which was caused by the
fact that we used a trivial semiring with \zero~$=$~\one. One can think of this
case as \hdia~being a monoid united with itself. The case of edge-labelled
graphs is more interesting. There we had as many operations as there were
elements in the edge label semiring~\hcode{s}. It turns out that every operation
$\xrightarrow{x}$ is united with \hdia. In other words, edge-labelled graphs
form a \emph{semiring of united monoids}.

@@

\begin{itemize}
    \item $(\mathbb{Z}^{\ge 0}, \max, 0)$: the execution time of programs $a$
    and $b$ composed in parallel is defined to be the maximum of their execution
    times:
    \[
    \text{\hcode{time}}(a\ \text{\code{|}}\ b) = \max(\text{\hcode{time}}(a),\ \text{\hcode{time}}(b))
    \]

    \item $(\mathbb{Z}^{\ge 0}, +, 0)$: the execution time of programs $a$ and
    $b$ composed in sequence is defined to be the sum of their execution times:
    \[
    \text{\hcode{time}}(a\ \text{\code{;}}\ b) = \text{\hcode{time}}(a) + \text{\hcode{time}}(b)
    \]
\end{itemize}

@@

\noindent
Interestingly, parallel and sequential composition of programs also forms a
united monoid, where the empty program is the unit. One can therefore call the
cost model function $\text{\hcode{time}}$ a \emph{united monoid
homomorphism\footnote{
    A \emph{homomorphism} is a structure-preserving map between two algebraic
    structures.
}}. A related example can be found in~\cite{beaumont2017concepts}, where
asynchronous circuit specifications are composed in parallel and in sequence,
using graphs labelled with the semiring of Boolean predicates.

% Naturally ordered
% Use trees/united monoids for modelling parallel/sequential composition
% What about tries?
% Equivalence classes?
% Languages

\section{Related work}\label{sec-related-work}

This paper continues the research on \emph{algebraic graphs} started
in~\cite{mokhov_alga}. Since their introduction, algebraic graphs have been
implemented in several languages (including Agda, Fsharp, Haskell, PureScript,
Scala, TypeScript, and R) and found applications in industry, e.g., in GitHub's
static program analysis library Semantic~\cite{semantic_paper}\cite{semantic_repo}.
In this paper we further distill the essence of the algebraic approach to
working with graphs by replacing the data type

\begin{lstlisting}[basicstyle=\sffamily\smaller]
data Graph a = Empty | Vertex a | Overlay (Graph a) (Graph a) | Connect (Graph a) (Graph a)
\end{lstlisting}

\noindent
with binary trees parameterised by a semiring. We remove the \hcode{Empty}
constructor to make it possible to distinguish non-empty structures at the
type level and also to keep expressions free from redundant empty leaves. By
merging the \hcode{Overlay} and \hcode{Connect} constructors into a single
\hcode{Node} parameterised by a semiring, we get a more general data structure.
This allows us to resolve one of the main limitations of the work
in~\cite{mokhov_alga}, which was the lack of support for edge labels.

Two other popular approaches for representing graphs in functional programming
languages were developed by King and Launchbury~\cite{1995_king_graphs}
(based on adjacency lists, available from the \code{containers} library)
and Erwig~\cite{2001_erwig_inductive} (based on \emph{inductive graphs}, where
graphs can be decomposed into a \emph{context}, i.e., a vertex with its
neighbourhood, and the rest of the graph). These and other classic graph
representations can be obtained from our tree expressions via folding. Note
that, as we remarked in~\S\ref{sec-simplicial-set}, folding a tree expression
into a ``flat'' graph representation too early can be undesirable because of the
loss of compactness and the need to maintain various correctness invariants.

% Add a note to the section on graphs about the invariant

Using semirings as a general framework for solving problems on graphs is, of
course, not new. Dolan's ``Fun with Semirings''~\cite{2013_semirings_dolan} is a
notable example of parameterising matrices with semirings, which motivated the
author of this paper to apply semirings in the context of algebraic graphs. In
a more recent work, Kidney and Wu~\cite{weighted_search} use semirings for
generalising algorithms for weighted search on graphs represented by functions
from vertices to their neighbours.

% Brent's sparks: https://byorgey.wordpress.com/2018/10/01/monoidal-sparks/
% Two monoid products: direct and sparked are monoids with the same identity

Jeremy Gibbons defines operations \hcode{above} and \hcode{beside} for composing
directed acyclic graphs~\cite{1995_gibbons_algebra}. They have the same unit
(the empty graph) but the operation \hcode{beside} is defined only for graphs of
matching types. One can therefore think of \hcode{above} and \hcode{beside} as
\emph{partial united monoids} or \emph{united categories}. Similarly, Santos and
Oliveira~\cite{typed_matrices} show how to compose type-compatible matrices
vertically and horizontally using the operations \hcode{fork} and \hcode{join},
respectively, whose common unit is the empty matrix.

Various flavours of parallel and sequential composition often form united
monoids. In the paper on Concurrent Kleene Algebra~\cite{hoare2011concurrent},
Hoare et al. use the term ``bimonoid'' to refer to such structures (without
investigating them in detail). In category theory, ``bimonoid'' is used for an
unrelated concept (a structure with a monoid and a
comonoid~\cite{porst_bimonoids}), hence our decision to use the term
``united monoids'', which appears to be unused and is also more specific
compared to ``bimonoid''.

Rivas and Jaskelioff discuss various notions of computation viewing them as
monoids in the category of endofunctors~\cite{rivas_jaskelioff_2017}. Two of
these monoids, namely \hcode{Applicative} and \hcode{Monad}, turn out to have
the same unit \hcode{pure = return}, and they can therefore be considered
\emph{united monoids in the category of endofunctors}. Investigating categorical
equivalents of the presented ideas is an interesting direction of future work.
